---
title: "How to use the geneticae package"
subtitle: "Statistical Tools for the Analysis of Multi Environment Agronomic Trials"
author: |
    | Julia Angelini
    | Marcos Prunello
    | Gerardo Cervigni
    |
    | Centro de Estudios Fotosintéticos y Bioquímicos
    | Universidad Nacional de Rosario
    | Rosario, Argentina
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use the geneticae package}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(warning = F, message = F, out.width = "60%")
```

<style type="text/css">

body{
text-align: justify}
</style>
  
  
<!-- devtools::build_vignettes() -->
# Getting started

<!-- __Installing the package.__ To install the released version of `geneticae` from -->
<!-- [CRAN](https://CRAN.R-project.org): -->

<!-- ``` {r, eval=F} -->
<!-- install.packages("geneticae") -->
<!-- ``` -->

You can install the development version from our [GitHub repo](https://github.com/jangelini/geneticae) with:

``` {r, eval=F}
# install.packages("devtools")
devtools::install_github("jangelini/geneticae")
```



__Loading the package.__ Once the `geneticae` package is installed, it needs to be loaded by:

```{r}
library(geneticae)
```

__Help files.__  Detailed information on `geneticae` package functions can be obtained from  help files using `help(package="geneticae")`.
The help file for a function, for example `imputation` can be obtained using `?imputation` or `help(imputation)`.


# Introduction

Understanding the relationship between crops performance and environment is a
key problem for plant breeders and geneticists. In advanced stages of breeding
programs, in which few genotypes are evaluated, multi-environmental trials (MET)
are one of the most used experiments. Such studies test a number of genotypes in
multiple environments in order to identify the superior genotypes according to
their performance. In these experimentes, crop performance is modeled as a
function of genotype (G), environment (E) and genotype-environment interaction
(GEI). The presence of GEI generates differential genotypic responses in the
different environments (Crossa et al., 1990; Cruz Medina, 1992; Kang and Magari,
1996). Therefore appropriate statistical methods should be used to obtain an
adequate GEI analysis, which is essential for plant breeders (Giauffret et al.,
2000).

The average performance of genotypes through different environments can only be
considered in the absence of GEI (Yan and Kang, 2003). However, GEI is almost
always present and the comparison of the mean performance between genotypes is
not enough. The most widely used methods to analyze MET data are based on
regression models, analysis of variance (ANOVA) and multivariate techniques. In
particular, two statistical models are widely used among plant breeders as they
provide useful graphical tools for the study of GEI: the Additive Main effects
and Multiplicative Interaction model (AMMI) (Kempton, 1984; Gauch, 1988) and the
Site Regression Model (SREG) (Cornelius et al., 1996; Gauch and Zobel, 1997).
However, these models are not always efficient enough to analyze MET data
structure of plant breeding programs. They present serious limitations in the
presence of atypical observations and missing values, which occur very
frequently. To overcome this, several imputation alternatives and a robust AMMI
were recently proposed in literature.

Although there are a few R packages which tackle different aspects of the
analysis of MET data, there aren't any packages capable of performing all the
steps that need to be considered. The `geneticae` package was created to gather
in one place the most useful functions for this type of analysis and it also
implements new methodology which can be found in recent literature. More
importantly, `geneticae` **is the first package to implement the robust AMMI
model and implements new imputation methods not available before**. In addition,
there is no need to preprocess the data to use the `geneticae` package, as it
can handle different dataset formats. Genotypes, environments, repetitions (if
any) and phenotypic traits of interest can be presented in any order and there
is no restriction on columns names, as it is the case with previous published
packages. Also, extra information that will not be used in the analysis may be
present in the datasets. Finally, `geneticae` offers a wide variety of options
to customize the biplots, which are part of the graphical output of these
methods.

# Datasets

The `geneticae` package provides two datasets to illustrate the methodology
included to analize MET data.

* `yan.winterwheat` dataset: yield of 18 winter wheat varieties grown in nine
environments in Ontario at 1993. Although four blocks or replicas in each
environment were performed in the experiment, only yield mean for each variety
and environment combination was available in the dataset obtained from the
[agridat](https://CRAN.R-project.org/package=agridat) package (Wright, 2018).


  ```{r}
  data(yan.winterwheat)
  head(yan.winterwheat)
  ```

* `plrv` dataset: resistance study to PLRV (Patato Leaf Roll Virus) causing leaf
curl. 28 genotypes were experimented at 6 locations in Peru. Each clone was
evaluated three times in each environment, and yield, plant weight and plot were
registered. This dataset was also obtained from the
[agricolae](https://CRAN.R-project.org/package=agricolae) package (de Mendiburu,
2020).


  ```{r}
  data(plrv)
  head(plrv)
  ```


# Statistical models for multi-environmental trials

## AMMI model

The *AMMI model* (Gauch, 1988, 1992) is widely used to analyse the effect of the
GEI. This model explores a matrix of GEI by performing a singular value
decomposition (SVD) on the residuals matrix from a one-way ANOVA with fixed
effects for environments and genotypes. The result of the first two
multiplicative terms of SVD is often presented in a biplot called GE and
represents a two-rank approximation of GE effects.

The `rAMMI()` function returns the GE biplot. Data in long format is required by
this function, i.e. each row corresponds to one observation and each column to
one variable (genotype, environment, repetition (if any) and the observed
phenotype). If each genotype has been evaluated more than once at each
environment, the phenotypic mean for each combination of genotype and
environment is internally calculated and then the model is estimated. Extra
variables that will not be used in the analysis may be present in the dataset.
Missing values are not allowed (but can be imputated, see below).

The GE biplot for `yan.winterwheat` dataset is shown in Figure 1.

```{r, fig.align='center', fig.cap='Figure 1: GE biplot.'}
rAMMI(yan.winterwheat, genotype = "gen", environment = "env", 
      response = "yield", type = "AMMI", footnote = F, titles = F)
```

The AMMI model, in its standard form, assumes equal weights for all the entries
of two-way dataset and that no outliers are present in the data. To overcome the
problem of data contamination with outlying observations, Rodrigues et al.
(2015) proposed five robust AMMI models, which can be obtained in two stages:
(i) fitting a robust regression model with an M-Huber estimator (Huber, 1981) to
replace the ANOVA model; and (ii) using a robust SVD or principal components
analysis (PCA) procedure to replace the standard SVD. Until now, robust AMMI
models  were not available in any R package. All robust biplots proposed by
Rodrigues et al. (2015) can be obtained using `rAMMI()` function. The argument
`type` can be used to specify the type of modelo to be fitted (`"rAMMI"`,
`"hAMMI"`, `"gAMMI"`, `"lAMMI"` or `"ppAMMI"`). Since the sample
`yan.winterwheat` dataset does not present outliers, the conclusions obtained
with robust biplots will not differ from those made with the classical biplot
(Rodrigues, et al. 2015). Thus, no interpretation of the robust biplots is
presented in this tutorial.


## Site Regression model

The *Site Regression model* (SREG, also called *genotype plus
genotype-by-environment model* or *GGE model*) is another powerful tool for the
analysis and interpretation of MET data in breeding programs. GGE explores
genotype (G) and EGI effects performing a SVD on the residuals matrix from the
one-way ANOVA model with fixed effects for environments. To select cultivars the
effect of G and GEI must be considered simultaneously so SREG model is superior
to AMMI for the visualization of patterns in MET data.

The `GGEmodel()` function is a wrapper for the `GGEModel()` function of the
[GGEBiplots](https://CRAN.R-project.org/package=GGEBiplots) package (Dumble,
2017) which allows for the presence of repetitions or extra variables in the
dataset. All the combinations between genotypes and environments must be
present.

<!-- Since `rAMMI()` requires data in long format, repetitions and extra variables not used in the analysis may be present in the dataset are allowed, and all combination of genotypes and environments must be evaluated. -->

```{r}
GGE1 <- GGEmodel(yan.winterwheat, genotype = "gen", environment = "env", 
                 response = "yield", centering = "tester")
```


The output from `GGEmodel()` is a list with the following elements:

* `coordgenotype`: plot coordinates for all genotypes in each component. 
* `coordenviroment`: plot coordinates for all environments in each component.
* `eigenvalues`: vector of eigenvalues from each component.
* `vartotal`: overall variance.
* `varexpl`: percentage of variance explained by each component. 
* `labelgen`: genotype names.
* `labelenv`: environment names.
* `axes`: axis labels.
* `Data`: scaled and centered input data. 
* `centering`: centering method.
* `scaling`: scaling method.
* `SVP`: SVP method.

The result of the first two multiplicative terms of the SVD is often presented
in a GGE biplot (Yan et al., 2000), which represents a rank-two approximation of
the G + GEI effects. Plant breeders have found GGE biplots to be useful tools
for the analysis of mega environment experiments (Yan et al., 2001; Yan and
Rajcan, 2002) and genotype and environment evaluation (Bhan et al., 2005; Kang
et al., 2006; Yan et al., 2007). The GGE biplot addresses many issues relative
to genotype and test environment evaluation. Considering the average performance
of each genotype, this plot can be used to evaluate specific and general
adaptation. In addition, environments can be visually grouped according to their
ability to discriminate among genotypes and their representativeness of other
test environments. GGE biplot reveals the *which-won-where* pattern and allows
to recommend specific genotypes for each environment (Yan and Tinker, 2005).

Using the output from `GGEmodel()`, `GGEPlot()` builds several GGE biplots, in
which cultivars are shown in lower case and environments in upper case. The plot
also displays the methods used for centering, scaling and SVD. Optionally, the
percentage of G + GEI explained by the two axes can be added as a footnote with
`footnote = T`, as well as a tittle with `titles = T`. A basic biplot is
produced with the option `type="Biplot"` (Figure 2).

```{r, fig.align='center', fig.cap='Figure 2: basic biplot. 78% of the total variability of G + GE is explained.'}
GGEPlot(GGE1, type="Biplot", footnote = F, titles = F)
```

Breeders usually want to identify the most suitable cultivars for a particular
environment of interest. To do this with GGE biplots, Yan and Hunt (2002)
suggest making an axis of that environment, drawing a line that passes through
environment marker and biplot origin. This can be done setting `type="Selected
Environment"` and providing the name of the environment in `selectedE` (Figure
3).

```{r, fig.align='center', fig.cap='Figure 3: comparison of cultivare performance in a selected environment (OA93).'}
GGEPlot(GGE1, type="Selected Environment", selectedE = "OA93", 
        footnote = F, titles = F)
```

Another goal of plant breeders is to determine which is the most suitable
environment for a genotype Yan and Hunt (2002) suggest plotting a line that
passes through the origin and a cultivar of interest. Figure 4 illustrates how
to visualize the relative adaptation of that cultivar in different environments.
The argument  `type` produces this plot when set to `"Selected Genotype"`, while
`selectedG` can be used to select the genotype.

```{r,  fig.align='center', fig.cap='Figure 4: comparison of the performance of cultivar Luc in different environments.'}
GGEPlot(GGE1, type="Selected Genotype", selectedG = "Luc", 
        footnote = F, titles = F)
```

It is also possible to compare two cultivars linking them with a line and a
segment perpendicular to it  (Figure 5).

```{r,  fig.align='center', fig.cap='Figure 5: comparison of the cultivars Kat and Cas.'}
GGEPlot(GGE1, type = "Comparison of Genotype", 
        selectedG1 = "Kat", selectedG2 = "Cas", 
        footnote = F, titles = F, axis_expand = 1.5)
```

The polygonal view of the GGE biplot proposed by Yan (1999) provides an
effective way to visualize the *which-won-where* pattern of a MET dataset
(Figure 6). Cultivars in the vertices are those with the longest vectors, in
their respective directions, which is a measure of the ability to respond to
environments. The dotted lines perpendicular to the polygon sides divide the
biplot into mega environments, each of which has a vertex cultivar, which is the
one with the highest yield (phenotype) in all environments found in it.

```{r,  fig.align='center', fig.cap='Figure 6: polygon view of the GGE biplot, showing which cultivars presented highest yiel in each environment.'}
GGEPlot(GGE1, type="Which Won Where/What", footnote = F, titles = F, axis_expand = 1.5)

```

Selecting cultivars within each megaenvironments will be an issues among plant
breeders. According to Figure 6, __zav__ is the best cultivar for environments
in one of the megaenvironments and __fun__ in the other. However, breeders do
not select a single cultivar in each megaenvironment. Instead, they evaluate all
cultivars in order to get an idea of their performance (yield and stability).
The GGE biplot based on *genotype-focused scaling* can be used to visualize both
genotype mean performance and stability. This visualization is achieved drawing
an environment coordinate (AEC). For example, Figure 7 shows the AEC for the
megaenvironment composed of he environments "BH93", "EA93", "HW93", "ID93",
"NN93", "RN93", "WP93". The abscissa represents the G effect and the ordinate
GEI associated with each genotype, which is a measure of the variability or
instability of the genotype. A greater projection on AEC, regardless of
direction, means greater instability.

```{r,  fig.align='center', fig.cap='Figure 7: average environment view of the GGE biplot based on genotype-focused scaling, showing mean yield and stability of genotypes.'}
data <- yan.winterwheat[yan.winterwheat$env %in% c("BH93", "EA93","HW93", "ID93",
                                                   "NN93", "RN93", "WP93"), ]
GGE2 <- GGEmodel(data, genotype = "gen", environment = "env", 
                 response = "yield", SVP = "row")
GGEPlot(GGE2, type = "Mean vs. Stability", footnote = F, titles = F)
```

In the GGE biplot it is also possible to visualize mean yield and stability of
genotypes in yield units __per se__ (Figure 8). The ideal cultivar, with the
highest yield and absolute stability, is represented by a small circle. This
ideal genotype is used as a reference, as it rarely exists. The distance between
cultivars and the ideal one can be used as a measure of convenience. Concentric
circles help to visualize these distances. In the example, _Fun_ is the closest
one to the ideal crop, and therefore the most desirable one, followed by _cas_
and _hay_, which in turn are followed by _rum_, _ham_, _rub_, _zav_, _del_ and
_reb_, etc.


```{r,  fig.align='center', fig.cap='Figure 8: Classification of genotypes with respect to the ideal genotype.', warning=FALSE}
GGEPlot(GGE1, type = "Ranking Genotypes", footnote = F, titles = F)
```

Although METs are performed to study cultivars, they are equally useful for the
analysis of the environments. This includes several aspects: (i) evaluating
whether the target region belongs to one or more megaenvironments; (ii)
identifying better test environments; (iii) detecting redundant environments
that do not provide additional information on cultivars; and (iv) determining
environments that can be used for indirect selection.

In Figure 9 environments are connected to the origin through vectors, allowing
us to understand the interrelationships between them. The coefficient of
correlation between two environments it is approximated by the cosine of the
angle formed by the respective vectors. For example, the angle between the
vectors for the environments NN93 and WP93 is approximately 10$º$; therefore,
they are closely related; while RN93 and OA93 present a weak negative
correlation since the angle is slightly greater than 90$º$. The cosine of the
angles does not translate precisely into coefficients of correlation, since the
biplot does not explain all the variability in the dataset. However, they are
informative enough to understand the interrelationship between test
environments.

```{r,  fig.align='center', fig.cap='Figure 9: Relationship between environments.'}
GGE3 <- GGEmodel(data, genotype = "gen", environment = "env", 
                 response = "yield", SVP = "column")
GGEPlot(GGE3, type = "Relationship Among Environments", footnote = F, titles = F)
```

The biplot shown in Figure 10 helps to identify redundant environments. If some
environments have small angles, genotype information obtained from these
environments should be similar. If this similarity is always present through the
years, these environments are redundant and therefore it should be enough to
keeo only one of the in the experimentation. Obtaining the same or better
information using fewer environments would reduce cost and increase production
efficiency.

```{r,  fig.align='center', fig.cap='Figure 10: classification of environments with respect to the ideal environment.'}
GGEPlot(GGE1, type = "Ranking Environments", footnote = F, titles = F)
```

__FALTA DESCRIPCION AGREGAR!!!__

```{r,  fig.align='center'}
GGEPlot(GGE1, type="Discrimination vs. representativeness", footnote = F, titles = F)
```


# Imputation methods

One major limitation of the AMMI and SREG models is that they require a complete
two-way data table. Although METs are designed so that all genotypes are
evaluated in all environments, missing values are very common due to measurement
errors or destruction of plants by animals, floods or harvest problems. In
addition, genotypes might be incorporated or discarded during the study because
of their promising or poor performance. The `imputation()` function includes
several methods to overcome the problem of missing data, some of which  have
been recently published and were not available in any R package until now.

To present an example, some observations from the complete `yan.winterwheat` are deleted:

```{r}
# Generating missing data
yan.winterwheat[1,3] <- NA
yan.winterwheat[3,3] <- NA
yan.winterwheat[2,3] <- NA
```

Imputation of missing values with the `"EM-AMMI"` method can be performed as follows:

```{r}
imputation(yan.winterwheat, nPC = 2, genotype = "gen", environment = "env", 
           response = "yield", type = "EM-AMMI")
```

The other methods available in `geneticae` are: `"EM-SVD"`, `"Gabriel"`, `"WGabriel"` and `"EM-PCA"`.
 

## References

Giauffret, C., Lothrop, J., Dorvillez, D., Gouesnard, B., Derieux, M., 2000. Genotype x environment interactions in maize hybrids from temperate or highland tropical origin. Crop Sci. 40, 1004-1012. \n

Sam Dumble (2017). GGEBiplots: GGE Biplots with 'ggplot2'. R package version 0.1.1. <https://CRAN.R-project.org/package=GGEBiplots> \n

Felipe de Mendiburu (2020). agricolae: Statistical Procedures for Agricultural Research. R package version 1.3-2. <https://CRAN.R-project.org/package=agricolae>

## Sesssion Info

```{r, echo = FALSE}
sessionInfo()
```
