---
title: "Geneticae"
subtitle: "An R package to analyze data from agricultural research"
author: |
    | Julia Angelini
    | Marcos Prunello
    | Gerardo Cervigni
    |
    | Centro de Estudios Fotosintéticos y Bioquímicos
    | Universidad Nacional de Rosario
    | Suipacha 501
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geneticae}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style type="text/css">

body{
  text-align: justify}
  </style>
  
  
<!-- devtools::build_vignettes() -->
## Getting started

__Installing the package.__ To install the released version of _geneticae_ from
[CRAN](https://CRAN.R-project.org):

``` {r, eval=F, tidy=TRUE}
install.packages("geneticae")
```

And the development version from [GitHub](https://github.com/):

``` {r, eval=F, tidy=TRUE}
# install.packages("devtools")
devtools::install_github("jangelini/geneticae")
```



__Loading the package.__ Once geneticae package is installed, it needs to be made accessible to the current R session by `library(geneticae)`.



__Help files.__  Detailed information on `geneticae` package functions can be obtained from the help files using `help(package="geneticae")`.
The help file for a function, for example `imputation`, in a R session can be obtained using `?imputation` or `help(imputation)`.


## Introduction

Understanding the relationship between crops performance and environment is a key problem for plant breeders and geneticists. In advanced stages of breeding programs, where few genotypes are evaluated, multi-environmental trials (MET) is one of the most common experiments. They are conducted by testing a number of genotypes across multiple environments, allowing the identification of superior genotypes. Crop performance is a function of genotype (G), environment (E) and genotype x environment interaction (GEI). METs are essential due to the presence of GEI which generates differential genotypic responses in the different environments evaluated (Crossa et al., 1990; Cruz Medina, 1992; Kang and Magari, 1996). This is a particular problem for plant breeders (Giauffret et al., 2000), therefore appropriate statistical methods should be used to obtain an adequate GEI analysis.

_geneticae_ package provides tools to analize data from advanced stages of
breeding programs, where few genotypes are evaluated. The classic AMMI and SREG biplot can be obtained from the package, as well as a robust version of AMMI which overcome the fragility of the conventional method in presence of outliers. Finally, since these biplots can be obtained only if all combinations between environment and genotypes was recorded, imputation techniques were included.

Unlike the other existing functions that allow adjusting these models and obtaining their biplots, the included ones are less restrictive in the format of the input data set. The information of the genotypes, environments, repetitions (if any) and the phenotypic trait of interest can be in any order and there is no restriction on the name that these columns should receive. In addition, other variables that will not used in the analysis may be present in the dataset. Finally, the biplots created by the package can be customized by the user and as they are ggplot objects, they can be customized even more than the function allows using various functions of the ggplot2 package.



## Data set in geneticae

The _geneticae_ package provides two datasets to illustrate the methodology included to analize MET data. 

* `yan.winterwheat` dataset: yield of 18 winter wheat varieties grown in nine environments in Ontario in 1993. Although the experiment had four blocks or replicas in each environment, in the data set only the average yield for each variety and environment combination was is available. This dataset was obtained from _agridat_ package (Wright, 2018).


```{r, tidy=TRUE}
library(geneticae)
data(yan.winterwheat)
head(yan.winterwheat)
```

* `plrv` dataset: resistance study to PLRV (Patato Leaf Roll Virus) causing leaf curl. 28 genotypes were experimented at 6 locations in Peru. Each clone was evaluated three times in each environment, and the yield, plant weight and plot were registered. This dataset was obtained from _agricolae_ package (de Mendiburu, 2020).


```{r, tidy=TRUE}
data(plrv)
head(plrv)
```


## Running geneticae

### Site regression model

Site regression model also named genotype plus genotype-by-environment (GGE) model is a powerful tool for effective analysis and interpretation of MET data structure in breeding programs. GGE explore G + GE performing a singular value decomposition (SVD) on the residuals matrix obtained from a one-way ANOVA with fixed effects for E. 


`GGEmodel` function it is a modification of the GGEModel() function of the GGEBiplots (Dumble, 2017), being less restrictive in the format of the input dataset. The long format is required by this function, the observations must be the rows and the variables (genotypes, environments, repetitions (if any) and the observed phenotype) in the columns. If each genotype has been evaluated more than once in each environment, the function internally calculates the phenotypic mean for each combination of genotype and environment and then estimates the model. Other variables that will not be used in the model may be present in the database since the phenotypic trait of interest must be indicated in one of the arguments of the function. Missing values are not allowed, if there is an error message in R console will be appeared.


```{r, tidy=TRUE}
GGE1 <- GGEmodel(yan.winterwheat, genotype="gen",environment="env", response="yield", centering = "tester")
```


The output from `GGEmodel` function is a list with the following elements:

* `coordgenotype`: plotting coordinates for genotypes from all components. 
* `coordenviroment`: plotting coordinates for environments from all components.
* `eigenvalues`: vector of eigenvalues from each component.
* `vartotal`: overall variance.
* `varexpl`: percentage of variance explained by each component. 
* `labelgen`: genotype names.
* `labelenv`: environment names.
* `axes`: axis labels.
* `Data`: scaled and centered input data. 
* `centering`: centering method.
* `scaling`: scaling method.
* `SVP`: SVP method.



The result of the first two multiplicative terms of the SVD is often presented in a biplot called genotype plus genotype x environment interaction (GGE) (Yan et al., 2000). Such biplot represents a rank-two approximation of the G + GE effects. Increasingly, plant breeders have found GGE biplots as useful tools in mega environment analysis (Yan et al., 2001; Yan and Rajcan, 2002), and genotype and environment evaluation (Bhan et al., 2005; Kang et al., 2006; Yan et al., 2007). The GGE biplot visually addresses many issues relative to genotype and test environment evaluation. Considering the average genotypes performance, the GGE biplot allows to evaluate the specific and general adaptation of them. In addition, environments could be visually evaluated and grouped according to their ability to discriminate among genotypes and their representativeness of other test environments. GGE biplot reveals the which-won-where pattern and allows to recommend specific genotypes for each ME (Yan and Tinker, 2005). 


Using `GGEmodel()` information, `GGEPlot()` function builds the GGE biplots. In such biplots the cultivars are shown in lower case, to differentiate them from the environments, which are in capital letters. The first component is used as the abscissa and the second as ordinate. In addition, it is possible to add the centering, scale and SVD method used, and the percentage of G + GE explained by the two axes as a footnote with the argument `footnote = T`, as well as the tittle with `titles = T`.

Biplots that respond to specific objectives can be achieved. The basic biplot, in which the 78% of the total variability of G + GE is explained, can be obtained indicating `type="Biplot"` (Figure 1).

```{r,fig.height=4, fig.width=4, fig.align='center', fig.cap='Figure 1: basic biplot', tidy=TRUE}
GGEPlot(GGE1, type="Biplot")
```

Breeders in general are interested in identifying the most suitable cultivars for their area. To identify the best genotypes in an environment through the GGE biplot, Yan and Hunt (2002) suggest making an axis of the environment, for example OA93, drawing a line that passes through the identifier of the environment and the origin. Such biplot is obtained by indicating `type="Selected Environment"` and the name of the environment to examine `selectedE = "OA93"` (Figure 2).

```{r,fig.height=4, fig.width=4, fig.align='center', fig.cap='Figure 2: comparison of cultivare performance in a selected environment (OA93)', tidy=TRUE}
GGEPlot(GGE1, type="Selected Environment", selectedE = "OA93")
```


Another interest of plant breeders is to determine which is the most suitable environment for a cultivar. Figure 3 illustrates how to visualize the relative adaptation of the cultivar _Luc_ in different environments, `type="Selected Genotype"` and the name of the genotype to examine `selectedG = "Luc"`. Yan and Hunt (2002) suggest plotting a line that passes through the biplot origin and the Luc marker, which we will call the Luc axis.

```{r,fig.height=5.5, fig.width=5.5, fig.align='center', fig.cap='Figure 3: comparison of the performance of cultivar Luc in different environments', tidy=TRUE}
GGEPlot(GGE1, type="Selected Genotype", selectedG = "Luc")
```


To compare two cultivars, for example _Kat_ and _Cas_, a connector line that connect the genotypes to be compared and also a perpendicular to the connector line is drawn (Figure 4). 
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', fig.cap='Figure 4: comparison of two cultivars, Kat and Cas', tidy=TRUE}
GGEPlot(GGE1, type="Comparison of Genotype", selectedG1 = "Kat", selectedG2 = "Cas")
```

The polygonal view of the GGE biplot proposed by Yan (1999) provides an effective way to visualize the which-won-where  pattern of an EMA dataset (Figure 5). Vertex cultivars are those with the longest vectors, in their respective directions, which is a measure of the ability to respond to environments. The perpendicular lines on the sides of the polygon divide the biplot into megaenvironments, each of which has a vertex cultivar, the one with the highest yield in all the environments found in it.

```{r,fig.height=5.5, fig.width=5.5, fig.align='center', fig.cap='Figure 5: polygon view of the GGE biplot, showing which cultivar yielded best in which environments', tidy=TRUE}
GGEPlot(GGE1, type="Which Won Where/What")
```




* Relationship Among Environments, is obtained by indicating `type="Relationship Among Environments"`.
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
GGEPlot(GGE1, type="Relationship Among Environments")
```



* Evaluating of environments based on both discriminating ability and representativeness, indicating `type="Discrimination vs. representativeness"`.
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
GGEPlot(GGE1, type="Discrimination vs. representativeness")
```

* Ranking of environments with respect to the ideal environment, indicating `type="Ranking Environments"`.
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
GGEPlot(GGE1, type="Ranking Environments")
```

* Ranking of genotypes with respect to the ideal genotype, indicating `type="Ranking Genotypes"`.
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
GGEPlot(GGE1, type="Ranking Genotypes")
```

* Evaluating of cultivars based on both average yield and stability, indicating `type="Mean vs. Stability"`.

```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
GGEPlot(GGE1, type="Mean vs. Stability")
```

### AMMI

#### The classic AMMI model
The AMMI model (Gauch, 1988, 1992) is another widely used tools to
analyse and structure GE. This model is fit in two stages: first, the main effects (G and E) are estimated using ANOVA by least squares; second, the SVD is applied to the residuals from the ANOVA, i.e. to the interaction, to obtain the estimates for the multiplicative terms of the AMMI model. In order to visualize the interaction patterns, the Genotipe-Environment (GE) biplots are constructed.

To run `rAMMI` function, as in the `GGEmodel` function, a dataset with genotype, environment, repetitions (if there is), and the response variable must be provided. The name of the columns that contains genotype, environment, repetitions and the response must be indicated. The output is a biplot.

The biplots from classic AMMI model, using `type="AMMI"`, that can be produced using _yan.winterwheat_ dataset is shown below.

```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
rAMMI(yan.winterwheat,genotype="gen",environment="env", response="yield", type = "AMMI")
```

#### Robust AMMI model
The AMMI model, in its standard form, implicitly assumes equal weights for all entries of the two-way dataset and that no outliers are present in the data.
To overcome the problem of data contamination with outlying observations, Rodrigues et al. (2015) propose five robust AMMI models, which can be obtained in two stages as follows: (i) use the robust regression based on the M-Huber estimator (Huber, 1981) to replace the ANOVA model; and (ii) use a robust SVD/PCA procedure to replace the
standard SVD.

The biplots of the five robust AMMI model proposed, using _yan.winterwheat_ dataset are shown below.

* `type = "rAMMI"`
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
rAMMI(yan.winterwheat, genotype="gen",environment="env", response="yield", type = "rAMMI")
```

* `type = "hAMMI"`
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
rAMMI(yan.winterwheat, genotype="gen",environment="env", response="yield", type = "hAMMI")
```

* `type = "gAMMI"`
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
rAMMI(yan.winterwheat, genotype="gen",environment="env", response="yield", type = "gAMMI")
```

* `type = "lAMMI"`
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
rAMMI(yan.winterwheat, genotype="gen",environment="env", response="yield", type = "lAMMI")
```

* `type = "ppAMMI"`
```{r,fig.height=5.5, fig.width=5.5, fig.align='center', tidy=TRUE}
rAMMI(yan.winterwheat, genotype="gen",environment="env", response="yield", type = "ppAMMI")
```

If repetitions are available in dataset, as _plrv_, the name of the column that contains the replications must be indicated.

### Imputation methods

One major limitation of the included models is that require a complete two-way data table. Although METs are designed so that all genotypes are evaluated in all environments, the presence of missing values is very common. These occur for example due to the incorporation of new genotypes, measurement errors or natural causes such as, for example, destruction of plants by animals, floods or during harvest.

Numerous methodologies have been proposed in order to overcome the lack of balance problem. The _yan.winterwheat_ dataset was used as example. Since the dataset was complete, some observations are eliminated in order to show the imputation methodologies.

```{r, tidy=TRUE}
# generates missing data
yan.winterwheat[1,3]<-NA
yan.winterwheat[3,3]<-NA
yan.winterwheat[2,3]<-NA
```

* __GabrielEigein__ proposed by Arciniegas-Alarcón S., et al. (2010).

```{r, tidy=TRUE}
imputation(yan.winterwheat, PC.nb=2, genotype="gen",environment="env", response="yield", type="EM-AMMI")
```

* __EM-AMMI__ proposed by Gauch and Zobel (1990).

```{r, tidy=TRUE}
imputation(yan.winterwheat, PC.nb=1, genotype="gen",environment="env", response="yield", type="EM-AMMI")
```

* __EM-SVD__ proposed by Perry (2009)

```{r, tidy=TRUE}
imputation(yan.winterwheat, genotype="gen",environment="env", response="yield", type="EM-SVD")
```

* __WGabriel__ proposed by Alarcon.....

```{r, tidy=TRUE}
imputation(yan.winterwheat, genotype="gen",environment="env", response="yield", type="WGabriel")
```

* __EM-PCA__ proposed by 

```{r, tidy=TRUE}
imputation(yan.winterwheat, genotype="gen",environment="env", response="yield", type="EM-PCA")
```

If repetitions are available in dataset, as _plrv_, the name of the column that contains the replications must be indicated.

## References

Giauffret, C., Lothrop, J., Dorvillez, D., Gouesnard, B., Derieux, M., 2000. Genotype x environment interactions in maize hybrids from temperate or highland tropical origin. Crop Sci. 40, 1004-1012. \n

 Sam Dumble (2017). GGEBiplots: GGE Biplots with 'ggplot2'. R package version 0.1.1. https://CRAN.R-project.org/package=GGEBiplots \n

Felipe de Mendiburu (2020). agricolae: Statistical Procedures for Agricultural Research. R package version 1.3-2. https://CRAN.R-project.org/package=agricolae

## Sesssion Info

```{r, echo = FALSE}
sessionInfo()
```
