---
title: "Geneticae"
subtitle: "An R package to analyze data from agricultural research"
author: |
    | Julia Angelini
    | Marcos Prunello
    | Gerardo Cervigni
    |
    | Centro de estudios fotosinteticos y bioquimicos
    | Universidad Nacional de Rosario
    | Suipacha 501
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geneticae}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


##1. Getting started

The main program of R should be already installed in the platform of your computer (Windows, Linux or MAC). If it is not installed yet, you can download it from the R project (\link{www.r-project.org}) of a repository CRAN, R Core Team (2018).


__Installing the package.__ To install the _Geneticae_ package:

```{r,eval=FALSE}
install.packages("devtools")
library(devtools)
install_github("jangelini/geneticae")
```


__Loading the package.__ Once geneticae package is installed, it needs to be made accessible to the current R session by the command. To load the `Geneticae` package in your R session, type `library(Geneticae)`.



__Help files.__  Detailed information on `Geneticae` package functions can be obtained in the help files obtained using `help(package="agricolae")`. The help file of particular command for the function `imputation` in a R session, can be obtained using `?imputation` or help(imputation).


For a complete functionality, geneticae requires other packages
\begin{itemize}
\item HOLA
\end{itemize}


##2. Introduction


The geneticae package offers functions for the analysis of genotype interaction by environment, an essential factor in an improvement program.





## 8. Sesssion Info

```{r, tidy=TRUE, echo = FALSE}
sessionInfo()
```
